Canonical rings of rational divisors on projective space and Hirzebruch surfaces

### 2015-07-12 surfaces-write-up/surface-write-up.tex

\DeclareMathOperator\di{Div}
\newcommand\bida{a}
\newcommand\bidb{b}
\newcommand\hirz{\mathbb{F}}

@@

\todo{macros: degrees}

@@

\begin{prop}
\label{prop:cone-generation}
Let $n,t \in \bz,$ let $\alpha_1, \ldots, \alpha_n \in \bq$ and let $k_i^j \in \bz$ with $1 \leq i \leq n, 1 \leq j \leq t.$ Define
\begin{align*}
	\Sigma = \{(d,c_1, \ldots, c_n) \in \bz^{n+1} : c_i \geq - d \alpha_i,1 \leq i \leq n \text{ and } \sum_{i=1}^{n}k_i^j = 0, 1 \leq j \leq t\}
\end{align*}
Let $e_1, \ldots, e_r \in \Sigma,$ with $e_i = (\delta^i, c_1^i, \ldots, c_n^i)$ be a set of {\bf extremal rays} of $\Sigma$, meaning that the $\Sigma$ is contained in the $R_{\geq 0}$ span of $e_1, \ldots, e_r$.  Then, as a semigroup, $\Sigma$ is generated by elements whose first coordinate is less than $\sum_{i=1}^{n}\delta_i$. Furthermore, every element $\sigma \in \Sigma$ can be written in a canonical form $\lambda + \sum_{i=1}^{r} w_i e_i,$ with the first coordinate of $\lambda$ less than $\sum_{i=1}^{r}\delta_i$ and $w_1,\ldots, w_r \in \bz$.

@@

\begin{proof}
By assumption, $\sigma \in \Sigma$ as $\sigma = \sum_{i=1}^{r} \bida_i e_i$ with $\bida_i \in \br$. Let $\{a\} := a - \lfloor a \rfloor$ denote the fractional part of $a$. Let $\lambda = \sum_{i=1}^{r}\{\bida_i\}$. Hence, we can write $\sigma = \lambda + \sum_{i=1}^{r}\lfloor \bida_i \rfloor e_i.$ Hence, $\sigma$ lies in the $\bz_{\geq 0}$ span of $\lambda, e_1,\ldots, e_r$, which all have first coordinate less than $\sum_{i=1}^{r}\delta_i$. Therefore, $\Sigma$ is generated by elements whose first coordinate is less than $\sum_{i=1}^{r}\delta_i$.
\end{proof}

@@

\begin{lem}
\label{lem:bound-ker-chi}
Retaining the notation of Lemma ~\ref{lem:composite-map}, if $\Sigma$ has extremal rays $e_1,\ldots, e_r$ in degrees $d_1, \ldots, d_r$ then $\ker \chi$ has a system of generators in degrees less than $\sum_{i=1}^{r}d_r-1$.
\end{lem}
\begin{proof}
Since $e_1, \ldots, e_r$ are extremal rays, by Proposition ~\ref{prop:cone-generation}, every element $\sigma \in \Sigma$ can be written in a canonical form $\lambda + \sum_{i=1}^{r}e_i$. Now, letting $\lambda_0 := 0,\lambda_1, \ldots, \lambda_m$ be all elements of $\Sigma$ so that we can write $\lambda_i = \sum_{i=1}^{r}\bida_i e_i$ with $0 \leq \bida_i < 1$ for all $i$. Then, for any $1 \leq i \leq j \leq m,$ we can write $\lambda_i + \lambda_j$ in the above canonical form, yielding a relation in degree at most $\deg \lambda_i + \deg \lambda_j \leq 2 \cdot \left( \sum_{i=1}^{r}d_r -1 \right).$ Furthermore, these relations generate all relations, as one can apply a sequence of these relations to put any $\sigma \in \Sigma$ into canonical form.
\end{proof}

@@

\todo{possibly change this to only have two equations}
\begin{prop}
\label{prop:pm-span-and-basis}
The functions $u^d \cdot \prod_{i=1}^n f_i^{c_i}$ so that 
\begin{align}
\label{align:pm-span}
\sum_{i=0}^{n} c_i \cdot \bida_i \cdot f_i = 0 & \text{ and } &c_i \geq \lfloor \alpha_i d\rfloor	
\end{align}
form a spanning set for $H^0(\bp^m, dD)$ over $k$. Furthermore, functions 
of the form $u^d \cdot \prod_{i=1}^n f_i^{c_i} $ such that
\begin{align}
\label{align:pm-basis}
c_i = -\lfloor d\alpha_i \rfloor & \text{ for } & i > m
\end{align}
form a basis for $H^0(\bp^m, dD)$ over $k$.
\end{prop}
\begin{proof}
By definition of $H^0(\bp^m,dD)$, functions satisfying conditions 
~\eqref{align:pm-span} lie in $H^0(\bp^m,dD)$. To complete the proof, it suffices to check functions satisfying conditions ~\eqref{align:pm-basis} form a basis of $H^0(\bp^m,dD)$. Note that there $\binom{m+ \lfloor dD \rfloor }{m}$ functions satisfying condition ~\eqref{align:pm-basis}. However we know $h^0(\bp^m,dD) = \binom{m+ \lfloor dD \rfloor }{m},$ so it suffices to show that those functions satisfying condition ~\eqref{align:pm-basis} are independent. This follows from the assumption that $f_0,\ldots, f_n$ form a basis of degree 1 rational functions, and so degree $\lfloor d \deg D \rfloor $ monomials in $f_0,\ldots, f_n$ form a basis of degree $\lfloor d \deg D \rfloor $ rational functions.
\end{proof}
\todo{the following bound is the generation bound beginning from evan's paper}

@@

\begin{thm}
\label{thm:proj-one-point}
Let $X = \bp^m$ with $\{x_0, \ldots, x_m\}$ a coordinate system for
$\bp^m$. Let $D = \alpha H$, with $\alpha = \frac{a}{b} \in \bq$
and $H = V(x_k)$ a hyperplane of $\bp^m$, be a divisor on $\bp^m$.
Let
\[
	\left\lfloor \frac{c}{d} \right\rfloor = \frac{c_0}{d_0} <
	\frac{c_1}{d_1} < \ldots < \frac{c_r}{d_r} = \frac{c}{d}
\]

@@

\begin{proof}
We know $\{F_{0}^{\vec{v}} : \deg \vec{v} = \left\lfloor \frac{a}{b}
\right\rfloor \}$ are minimal generators in degree $1$. Now let $i >
0$. If $F_{d_i}^{\vec{v}}$ were not minimal for some $\vec{v}$ such
that $\deg \vec{v} = c_i$, then $F_{d_i}^{\vec{v}} = F_{s}^{\vec{w}}
F_{t}^{\vec{z}}$ for some $s, t \in \bn$ and $\vec{w}, \vec{z} \in
\bz_{\geq 0}^{m + 1}$. In particular, we see that 
\begin{align*}
	&d_i = s + t \\
	&c_i = \deg \vec{v} = \deg \vec{w} + \deg \vec{z}
\end{align*}

@@

Now we show that no further generators are necessary. As done in
the proof of the $\bp^1$ case by O'Dorney \cite[Theorem 6]
{dorney:canonical}, each lattice point $(d', c') \in \bz_{\geq 0}^2$
such that $c' \leq d' \alpha$ is a non-negative linear combination of
$\vec{x}_i$ and $\vec{x}_{i + 1}$ where $\vec{x}_i := (d_i, c_i) \in
\bz^2$. Consequently, any element $\frac{u^{d'} x^{\vec{v}}}
{x_k^{c'}} \in R_D$ is expressible as
\begin{align*}
	\frac{u^{d'} x^{\vec{v}}} {x_k^{c'}} = \frac{u^{d_i} x^{\vec{w}}}
	{x_k^{c_i}} \frac{u^{d_{i + 1}} x^{\vec{z}}} {x_k^{c_{i + 1}}}
\end{align*}

@@

To do this, we extend the method of O'Dorney \todo{cite}. We first
view the cone inside of $\mathbb{R}^{n+1}$ given by elements of the
form $(d, c_1, \ldots, c_n)$ satisfying $c_i \ge -\alpha_i d$ for
all $i$ and $\sum_{i = 1}^n c_i (\bida_i + \bidb_i) =0$. By O'Dorney, this
has extremal rays given by

@@

Let $\Delta^{n-1}$ be the simplex with vertices $e_1, \ldots e_m$.
We can then intersect the hyperplane $H$ given $\sum_{i=1}^m \bida_i X_i
= 0$ to get a subspace $S$. The extremal rays of $S$ are given
by the set of $e_i$ contained in $H$ together with the points of
intersection $e_{i,j}$ of $H$ and the line segment connecting $e_i,
e_j \not \in H$.

>>>>

\DeclareMathOperator\di{Div}
\newcommand\bida{a}
\newcommand\bidb{b}
\newcommand\pdeg{\delta}
\newcommand\hirz{\mathbb{F}}

@@

@@

\begin{prop}
\label{prop:cone-generation}
Let $n,t \in \bz,$ let $\alpha_1, \ldots, \alpha_n \in \bq$ and let $k_i^j \in \bz$ with $1 \leq i \leq n, 1 \leq j \leq t.$ Define
\begin{align*}
	\Sigma = \{(d, c_1, \ldots, c_n) \in \bz^{n+1} : c_i \geq - d \alpha_i,1 \leq i \leq n \text{ and } \sum_{i=1}^{n}k_i^j = 0, 1 \leq j \leq t\}
\end{align*}
Let $e_1, \ldots, e_r \in \Sigma,$ with $e_i = (\pdeg^i, c_1^i, \ldots, c_n^i)$ be a set of {\bf extremal rays} of $\Sigma$, meaning that the $\Sigma$ is contained in the $R_{\geq 0}$ span of $e_1, \ldots, e_r$.  Then, as a semigroup, $\Sigma$ is generated by elements whose first coordinate is less than $\sum_{i=1}^{n}\pdeg_i$. Furthermore, every element $\sigma \in \Sigma$ can be written in a canonical form $\lambda + \sum_{i=1}^{r} w_i e_i,$ with the first coordinate of $\lambda$ less than $\sum_{i=1}^{r}\pdeg_i$ and $w_1,\ldots, w_r \in \bz$.

@@

\begin{proof}
\todo{define $\pdeg$, $\bida$, and $\bidb$}
By assumption, $\sigma \in \Sigma$ as $\sigma = \sum_{i=1}^{r} \bida_i e_i$ with $\bida_i \in \br$. Let $\{a\} := a - \lfloor a \rfloor$ denote the fractional part of $a$. Let $\lambda = \sum_{i=1}^{r}\{\bida_i\}$. Hence, we can write $\sigma = \lambda + \sum_{i=1}^{r}\lfloor \bida_i \rfloor e_i.$ Hence, $\sigma$ lies in the $\bz_{\geq 0}$ span of $\lambda, e_1,\ldots, e_r$, which all have first coordinate less than $\sum_{i=1}^{r}\pdeg_i$. Therefore, $\Sigma$ is generated by elements whose first coordinate is less than $\sum_{i=1}^{r}\pdeg_i$.
\end{proof}

@@

\begin{lem}
\label{lem:bound-ker-chi}
Retaining the notation of Lemma ~\ref{lem:composite-map}, if $\Sigma$ has extremal rays $e_1,\ldots, e_r$ in degrees $d_1, \ldots, d_r$ then $\ker \chi$ has a system of generators in degrees less than $\sum_{i=1}^{r}d_r-1$.
\end{lem}
\begin{proof}
\todo{define $\bida$ and $\bidb$}
Since $e_1, \ldots, e_r$ are extremal rays, by Proposition ~\ref{prop:cone-generation}, every element $\sigma \in \Sigma$ can be written in a canonical form $\lambda + \sum_{i=1}^{r}e_i$. Now, letting $\lambda_0 := 0,\lambda_1, \ldots, \lambda_m$ be all elements of $\Sigma$ so that we can write $\lambda_i = \sum_{i=1}^{r}\bida_i e_i$ with $0 \leq \bida_i < 1$ for all $i$. Then, for any $1 \leq i \leq j \leq m,$ we can write $\lambda_i + \lambda_j$ in the above canonical form, yielding a relation in degree at most $\deg \lambda_i + \deg \lambda_j \leq 2 \cdot \left( \sum_{i=1}^{r}d_r -1 \right).$ Furthermore, these relations generate all relations, as one can apply a sequence of these relations to put any $\sigma \in \Sigma$ into canonical form.
\end{proof}

@@

\todo{possibly change this to only have two equations}
\begin{prop}
\label{prop:pm-span-and-basis}
The functions $u^d \cdot \prod_{i=1}^n f_i^{c_i}$ so that 
\begin{align}
\label{align:pm-span}
\sum_{i=0}^{n} c_i \cdot \bida_i \cdot f_i = 0 & \text{ and } &c_i \geq \lfloor \alpha_i d\rfloor	
\end{align}

\noindent
\todo{the above ``and'' needs to be placed better}
form a spanning set for $H^0(\bp^m, dD)$ over $k$. Furthermore, functions 
of the form $u^d \cdot \prod_{i=1}^n f_i^{c_i} $ such that
\begin{align}
\label{align:pm-basis}
c_i = -\lfloor d\alpha_i \rfloor & \text{ for } & i > m
\end{align}

\noindent
form a basis for $H^0(\bp^m, dD)$ over $k$.
\end{prop}

\begin{proof}
By definition of $H^0(\bp^m,dD)$, functions satisfying conditions 
~\eqref{align:pm-span} lie in $H^0(\bp^m,dD)$. To complete the proof, it suffices to check functions satisfying conditions ~\eqref{align:pm-basis} form a basis of $H^0(\bp^m,dD)$. Note that there $\binom{m+ \lfloor dD \rfloor }{m}$ functions satisfying condition ~\eqref{align:pm-basis}. However we know $h^0(\bp^m,dD) = \binom{m+ \lfloor dD \rfloor }{m},$ so it suffices to show that those functions satisfying condition ~\eqref{align:pm-basis} are independent. This follows from the assumption that $f_0,\ldots, f_n$ form a basis of degree 1 rational functions, and so degree $\lfloor d \deg D \rfloor $ monomials in $f_0,\ldots, f_n$ form a basis of degree $\lfloor d \deg D \rfloor $ rational functions.
\end{proof}
\todo{the following bound is the generation bound beginning from evan's paper}

@@

\begin{thm}
\label{thm:proj-one-point}
Let $X = \bp^m$ with $\{x_0, \ldots, x_m\}$ a coordinate system for
$\bp^m$. Let $D = \alpha H$, with $\alpha = \frac{p}{q} \in \bq$
and $H = V(x_k)$ a hyperplane of $\bp^m$, be a divisor on $\bp^m$.
Let
\[
	\left\lfloor \frac{p}{q} \right\rfloor = \frac{c_0}{d_0} <
	\frac{c_1}{d_1} < \ldots < \frac{c_r}{d_r} = \frac{p}{q}
\]

@@

\begin{proof}
We know $\{F_{0}^{\vec{v}} : \deg \vec{v} = \left\lfloor \frac{p}{q}
\right\rfloor \}$ are minimal generators in degree $1$. Now let $i >
0$. If $F_{d_i}^{\vec{v}}$ were not minimal for some $\vec{v}$ such
that $\deg \vec{v} = c_i$, then $F_{d_i}^{\vec{v}} = F_{s}^{\vec{w}}
F_{t}^{\vec{z}}$ for some $s, t \in \bn$ and $\vec{w}, \vec{z} \in
\bz_{\geq 0}^{m + 1}$. In particular, we see that 
\begin{align*}
	&d_i = s + t \\
	&c_i = \deg \vec{v} = \deg \vec{w} + \deg \vec{z}
\end{align*}

@@

Now we show that no further generators are necessary. As done in
the proof of the $\bp^1$ case by O'Dorney \cite[Theorem 6]
{dorney:canonical}, each lattice point $(\delta, \gamma) \in \bz_{\geq 0}^2$
such that $c \leq d \alpha$ is a non-negative linear combination of
$\vec{x}_i$ and $\vec{x}_{i + 1}$ where $\vec{x}_i := (d_i, c_i) \in
\bz^2$. Consequently, any element $\frac{u^{\delta} x^{\vec{v}}}
{x_k^{\gamma}} \in R_D$ is expressible as
\begin{align*}
	\frac{u^{\delta} x^{\vec{v}}} {x_k^{\gamma}} = \frac{u^{d_i} x^{\vec{w}}}
	{x_k^{c_i}} \frac{u^{d_{i + 1}} x^{\vec{z}}} {x_k^{c_{i + 1}}}
\end{align*}

@@

To do this, we extend the method of O'Dorney \todo{cite}. We first
view the cone inside of $\mathbb{R}^{n+1}$ given by elements of the
form $(d, c_1, \ldots, c_n)$ satisfying $c_i \ge -d \alpha_i$ for
all $i$ and $\sum_{i = 1}^n c_i (\bida_i + \bidb_i) =0$. By O'Dorney, this
has extremal rays given by

@@

Let $\Delta^{n-1}$ be the simplex with vertices $e_1, \ldots e_m$.
We can then intersect the hyperplane $H$ given $\sum_{i=1}^m \bida_i X_i
= 0$ to get a subspace $S$. The extremal rays of $S$ are given
by the set of $e_i$ contained in $H$ together with the points of
intersection $e_{i, j}$ of $H$ and the edges of the simplex
connecting the $e_i, e_j \not \in H$.
\todo{more detail above}
